Finite-size scaling of free energy in the dimer model on a hexagonal domain

### 2020-04-08 AffineLieAlgebras/MQFT2018/paper.tex
Accepted to TMPh

\section{Model definition}
\label{sec:model-definition}
The configurations of the dimer model are perfect matchings (sets of non-touching edges, covering
all the vertices) on some graph ${\cal G}$ with some choice of weights $\omega(e)$ on the edges. The
model is solvable on the bipartite graphs, i.e. the partition function can be computed if the
weights are introduced in such a way that for each face bounded by 0 mod 4 edges there is an odd
number of negative edge weights and each face with 2 mod 4 edges has an even number of of negative
edge weights. Then the signs of the edge weights form a so-called ``Kasteleyn orientation'' on graph,
the weighting is called ``Kasteleyn weighting'' \cite{kenyon2001dominos,kenyon2009lectures}.

@@

We can draw a rhombus on a dual lattice around each edge in the configuration. The picture of
``cubes in the corner'' presented in the Fig.~\ref{dhf} is obtained. Let us write on the top of each
uppermost cube the height of its column of cubes. Looking at this picture from the top, we obtain a
height function defined on the rectangular domain of the square lattice.

\begin{figure}[htbp]
\center{\scalebox{0.4}{\includegraphics{my-loz}}}
\caption{\label{dhf}A configuration of dimers on the hexagonal lattice and a corresponding picture
  of ``cubes in the corner''.}
\end{figure}

Let us define the sizes $M$, $N$, and $K$ of the sides of the hexagon.
The above description can be formalized by
%set of configurations of the model can be defined all possible ways to write
setting the non-negative numbers up to $K$ in the boxes of the rectangular $M\times N$ table so that a value in
each box is not greater than values in the adjacent upper and left boxes
\begin{equation}
  \label{eq:1}
  h_{ij}\leq h_{i-1,j},\quad h_{ij}\leq h_{i,j-1}.
\end{equation}

@@

MacMahon formula is obtained for the following definition of the Kasteleyn matrix. Embed the
hexagonal lattice in the complex plane $\mathbb{C}$ in such a way that some edges are parallel to
the real line with and corresponding vertices have coordinates with integer real and imaginary
parts. Then take
\begin{eqnarray}
  \label{eq:18}
  {\cal K}(w,b)=q^{\Re w+\Im w} \quad \mathrm{if}\quad \Im w=\Im b\\
  {\cal K}(w,b)=1 \quad \mathrm{if}\quad \Im w\neq\Im b.
\end{eqnarray}

@@

\section{Numeric checks}
\label{sec:numeric-checks}
To check our result \eqref{eq:26} numerically, we first need to evaluate the integral in $f_{3}$.
The function $e^{-z}\frac{\xi(z)-\xi(0)}{z}$ is smooth and but some care is required when taking the
limit $z\to 0$ numerically. Evaluating the integral numerically with Mathematica we get
\begin{equation}
  \label{eq:2}
  \int_{0}^{\infty}e^{-z}Q(z)\dz=\int_{0}^{\infty}e^{-z}\frac{\xi(z)-\xi(0)}{z} dz = -0.080842.%0.08084228607897521
\end{equation}

@@

>>>>

\section{Model definition}
\label{sec:model-definition}
The configurations of the dimer model are perfect matchings (sets of
non-touching edges, covering all the vertices) on some graph
${\cal G}$. Boltzmann weight of a configuration is given by the
product of edge weights $\omega(e)$ for some choice of edge weights. The
partition function is given by the sum of Boltzmann weights over all
configurations: $Z=\sum_{\mathrm{conf}}\prod_{e\in\mathrm{conf}}\omega(e).$

We consider dimer model on a hexagonal lattice. There are three
possible directions of the edges. First we consider the following
choice of Bolzmann weights. Take positive number $q$ and embed the
hexagonal lattice in the complex plane $\mathbb{C}$ in such a way that
some edges $e$ are parallel to the real line and corresponding
vertices $w,b$ have coordinates with half-integer imaginary part (See left pane of the Fig.~\ref{dhf}). Then take
\begin{equation}
  \label{eq:18}
  \begin{array}{lll}
 \omega(e_{w\to b})=q^{\Im w}&  \mathrm{if} & \Im w=\Im b\\
  \omega(e_{w\to b})=1 & \mathrm{if}& \Im w\neq\Im b.
  \end{array}
\end{equation}

In the section \ref{sec:box-infinite-height} we will consider the case when $q=q(\Re w)$ is an arbitrary smooth function of the real part of coordinate.

In general the model is solvable on the bipartite graphs, i.e.
the partition function can be computed if the weights are introduced
in such a way that for each face bounded by 0 mod 4 edges there is an
odd number of negative edge weights and each face with 2 mod 4 edges
has an even number of of negative edge weights. Then the signs of the
edge weights form a so-called ``Kasteleyn orientation'' on graph, the
weighting is called ``Kasteleyn weighting''
\cite{kenyon2001dominos,kenyon2009lectures}.

@@

We can draw a rhombus on a dual lattice around each edge in the configuration. The picture of
``cubes in the corner'' presented in the Fig.~\ref{dhf} is obtained. Let us write on the top of each
uppermost cube the height of its column of cubes. Looking at this picture from the top, we obtain a
height function that is defined on the rectangular domain of the square lattice.

\begin{figure}[htbp]
\center{\scalebox{0.4}{\includegraphics{my-loz}}}
\caption{\label{dhf}On the left: a configuration of dimers on the hexagonal lattice and the choice of coordinates for the definition of weights (\ref{eq:18}). On the right: corresponding picture
  of ``cubes in the corner'' for the height function $h$.}
\end{figure}

Let us define the sizes $M$, $N$, and $K$ of the sides of the hexagon.
The above description of height function can be formalized by
%set of configurations of the model can be defined all possible ways to write
setting the non-negative numbers up to $K$ in the boxes of the rectangular $M\times N$ table so that a value in
each box is not greater than values in the adjacent upper and left boxes
\begin{equation}
  \label{eq:1}
  h_{ij}\leq h_{i-1,j},\quad h_{ij}\leq h_{i,j-1}.
\end{equation}

@@

@@

\section{Numeric checks}
\label{sec:numeric-checks}
To check our result \eqref{eq:26} numerically, we first need to evaluate the integral in $f_{3}$.
The function $e^{-z}\frac{\xi(z)-\xi(0)}{z}$ is smooth and continuous, but some care is required when taking the
limit $z\to 0$ numerically. Evaluating the integral numerically with Mathematica we get
\begin{equation}
  \label{eq:2}
  \int_{0}^{\infty}e^{-z}Q(z)\dz=\int_{0}^{\infty}e^{-z}\frac{\xi(z)-\xi(0)}{z} dz = -0.080842.%0.08084228607897521
\end{equation}

@@

Since a hexagon with an infinite side is similar to an infinite strip,
it is natural to compare our result with the results for finite size
corrections to the scaling of free energy for dimer model on a square
lattice on a strip, cylinder and Möbius strip. These results were
recently nicely summarized in the paper \cite{izmailian2017finite} and
further investigated in the paper \cite{Sh_Izmailian_2019}. There the
following general expansion for the free energy on $M\times N$ square
lattice was presented ($S=M\times N$ here):
\begin{equation}
  \label{eq:32}
  f=f_{\mathrm{bulk}}+\frac{2}{M}f_{s1}+\frac{2}{N}f_{s2}
  +f_{\mathrm{corn}}\frac{\ln S}{S} +\frac{1}{S} (f_{\mathrm{univ}}+f_{\mathrm{nonuniv}})
\end{equation}
In particular, on a strip the bulk term is
$f_{\mathrm{bulk}}=\frac{G}{\pi}$, where $G$ is the Catalan constant,
surface free energy is nonzero, $f_{s1}=-\frac{1}{4}\ln(1+\sqrt{2})$ 
and the coefficient of $\frac{1}{S}$ is equal to $\frac{\pi}{24}$. In
the case of cylinder and  Möbius strip logarithmic term also appears.

Comparing our results (\ref{eq:6}), (\ref{eq:8}), (\ref{eq:9}),
(\ref{eq:11}) with these expressions, we see that behavior on
hexagonal lattice in the presence of limit shape is different. In
particular, bulk term $f_{0}$ depends on the geometric parameters $a$
and $b$, boundary is
tension free, so no boundary term $f_{1}$ is zero, and geometric
parameters of the domain appear again in the coefficient of
$\varepsilon^{2}$.
